---
title: "Distributional Consequences of Climate Policy"
author: "Deanna Elliott, Lauren Harris, Mia Guarnieri"
date: "5/7/2022"
output: 
  html_document:
    code_folding: hide
---

Last year, the Biden Administration assembled an Inter-agency Working Group (IWG) tasked with
updating the United States Government’s Social Cost of Carbon (SCC), which has not been com-
prehensively updated since 2010. The Administration has also called on government agencies to
address environmental justice, racism, and equity concerns when considering policies designed to
mitigate climate change.

While the Interagency Working Group develops a new SCC, the Biden Administration’s “interim”
value is $51 per metric ton of CO2. The electricity sector is the second largest source of greenhouse
gas emissions in the U.S. (after transportation). In this homework, you will consider the distributional consequences of imposing a household electricity tax based on the SCC to address the climate
change problem.

We recommend using R and writing functions to compute your answers wherever possible.
Use the following set of facts:
• Consumers can be separated into two income groups: “high” and “low.” The data set provides
price (in $) and quantity (in kWh) estimates of demand per month for the two groups. Run
linear regressions (with an intercept) to estimate the demand curves for “high” and “low”
income consumers.
• Initially, there is no tax on electricity consumption.
• The current electricity price (without any taxes) is $.10 per kWh.
• The marginal cost of producing a kWh of electricity is linear and has a price-intercept of 0

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(rootSolve)
library(here)
library(janitor)
library(calecopal)
library(stats)
library(kableExtra)
```

1. One kWh of electricity emits 0.85 pounds of CO2. Assuming that the interim SCC correctly reflects the total social cost of one metric ton of CO2, what is the marginal externality cost per kwH of electricity?

```{r, results = FALSE}
#convert pounds to metric tons and then multiply by cost
(0.85/2204.62)*51

#(lbs carbon/1 metric ton)*SCC

```

**Assuming the interim SCC correctly reflects the total social cost of one metric ton of CO2, the marginal cost per kwH of electricity is $0.0197 (1.97 cents)**

2. What is the aggregate monthly demand curve for electricity? What is the supply curve for electricity? What is the “benefit” to consumers under the status quo? What is the “benefit” to producers under the status quo? What is the environmental cost under the status quo?

```{r}
data <- read_csv(here("data", "HW3_data.csv")) %>% 
  select(-1) %>% 
  clean_names()

model_demand_l <- lm(price_cents ~ q_low_kwh, data=data)
model_demand_h <- lm(price_cents ~ q_high_kwh, data=data)

#demand function
demand <- function(p, model){
  q <- (p - model$coefficients[[1]])/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
}

#aggregate demand
demand_agg <- function(p){
  q <- demand(p, model_demand_l) + demand(p, model_demand_h)
  return(q)
}

#price vector
price = seq(0, 30, length.out = 100)

#applying the aggregate demand function to the price vector
Qagg <- map(price, demand_agg) %>% unlist()

#making the new dataframe with Qagg and price
agg_data<- tibble(Qagg = Qagg, price = price)

#making predicted dataframes with low and high models
l_pred <- tibble(price = price,
                 demand = demand(price, model_demand_l))

l_pred[l_pred == 0] <- NA

h_pred <- tibble(price = price,
                 demand = demand(price, model_demand_h))

#MPC line
qfm <- demand_agg(10)

MPC_slope <- 10/qfm

supply <- function(q){
  p <- MPC_slope * q
  return(p)
}

MPC <- tibble(price = supply(agg_data$Qagg),
              demand = agg_data$Qagg)

#plot
ggplot() +
  geom_line(agg_data, mapping = aes(x = Qagg, y = price, color = "Aggregate", lty = "Aggregate")) +
  geom_line(l_pred, mapping = aes(x = demand, y = price, color = "Low income", lty = "Low income")) +
  geom_line(h_pred, mapping = aes(x = demand, y = price, color = "High income", lty = "High income")) +
  geom_line(MPC, mapping = aes(x = demand, y = price, color = "MPC", lty = "MPC")) +
  geom_segment(aes(x = 0, y = 1.97, xend = 819713.75, yend = 1.97, color = "MEC", lty = "MEC")) +
  geom_line(MPC, mapping = aes(x = demand, y = price + 1.97, color = "MSC", lty = "MSC")) +
  scale_color_manual(name = "Legend", values = cal_palette("kelp1")) +
  scale_linetype_manual(name = "Legend", values = c("solid", "dashed", "dashed", "solid", "dotdash", "solid")) +
  labs(x = "Quantity (kWh)",
       y = "Price (cents)",
       color = "Demand Curve") + 
  theme_minimal()
```

```{r}
#consumer surplus
CS <- function(p, model){
  q <- demand(p, model)
  cs <- 0.5*(model$coefficients[[1]] - p)*q
  return(cs)
}

CS_low <- CS(10, model_demand_l)
CS_high <- CS(10, model_demand_h)

CS_agg <- CS_low + CS_high

#producer surplus

PS <- function(q){
  0.5 * supply(q) * q
}

PS_calc <- PS(qfm)

#environmental cost
env_cost <- 1.97 * qfm
```

**The benefit to consumers is \$52987.22. The benefit to producers is \$26835.97. The environmental cost under the status quo is $10573.37**


3. How is the current consumer benefit divided between “high” and “low” income consumers?

```{r}

high_CS <- CS_high/100
low_CS <- CS_low/100
```

**Currently, the benefit to high income consumers is \$44,874.79, and the benefit to low income consumers is $8,112.43.**

4. Derive the optimal electricity tax (in cents per kWh) using the interim SCC. Noting that recent research has shown the poor face a disproportionate share of the impacts from climate change, assume that the climate externality is borne entirely by the “low” income group. What would be the effects of this tax on:

(a) The amount of electricity produced and consumed
(b) The price of electricity

```{r}
#new demand function
demand_51tax <- function(p, model){
  q <- ((p + 1.97) - model$coefficients[[1]])/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
}

#factoring in tax for demand curves
l_pred_new <- tibble(price = price,
                 demand = demand_51tax(price, model_demand_l))

l_pred_new[l_pred_new == 0] <- NA

h_pred_new <- tibble(price = price,
                 demand = demand_51tax(price, model_demand_h))

#create a new Dagg
demand_agg_new <- function(p){
  q <- demand(p + 1.97 , model_demand_l) + demand(p + 1.97, model_demand_h)
  return(q)
}

#applying the aggregate demand function to the price vector
Qagg_new <- map(price, demand_agg_new) %>% unlist()

#making the new dataframe with Qagg and price
agg_data_new <- tibble(Qagg = Qagg_new, price = price)

ggplot() +
  geom_line(agg_data, mapping = aes(x = Qagg, y = price, color = "Aggregate", lty = "Aggregate")) +
  geom_line(l_pred, mapping = aes(x = demand, y = price, color = "Low income", lty = "Low income")) +
  geom_line(h_pred, mapping = aes(x = demand, y = price, color = "High income", lty = "High income")) +
  geom_line(l_pred_new, mapping = aes(x = demand, y = price, color = "Low income (new)", lty = "Low income (new)")) +
  geom_line(h_pred_new, mapping = aes(x = demand, y = price, color = "High income (new)", lty = "High income (new)")) +
  geom_line(MPC, mapping = aes(x = demand, y = price, color = "MPC", lty = "MPC")) +
  geom_segment(aes(x = 0, y = 1.97, xend = 819713.75, yend = 1.97, color = "MEC", lty = "MEC")) +
  geom_line(agg_data_new, mapping = aes(x = Qagg_new, y = price, color = "Aggregate (new)", lty = "Aggregate (new)")) +
  scale_color_manual(name = "Legend", values = cal_palette(name = "kelp1", n = 8, type = "continuous")) +
  scale_linetype_manual(name = "Legend", values = c("dotdash", "solid", "dashed", "solid", "dashed", "solid", "solid", "solid")) +
  labs(x = "Quantity (kWh)",
       y = "Price (cents)",
       color = "Demand Curve") + 
  theme_minimal()

#finding the intersect
fm_eq <- function(p){
  (demand(p + 1.97 , model_demand_l) + demand(p + 1.97, model_demand_h))-(p/MPC_slope)
  }
peq <- uniroot(fm_eq, c(0,750000))$root
qeq <- demand_agg_new(peq)
```

**Optimal tax is 1.97 cents. The amount of electricity that is produced is 500216.43 kWh, and the price per kWh is 11.29 cents.**

(c) Overall welfare of “high” income consumers

```{r}
#new demand function
demand_51tax <- function(p, model){
  q <- ((p + 1.97) - model$coefficients[[1]])/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
}

CS_51tax <- function(p, model){
  q <- demand_51tax(p, model)
  cs <- 0.5*(model$coefficients[[1]] - p)*q
  return(cs)
}

CSh_51 <- CS_51tax(peq, model_demand_h)/100
```

**The welfare of high income consumers is $43524.05, as they do not bear the environmental costs.**

(d) Overall welfare of “low” income consumers

```{r}
env_cost51 <- (1.97 * qeq)/100

#new demand function
demand_51tax <- function(p, model){
  q <- ((p + 1.97) - model$coefficients[[1]])/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
}

CS_51tax <- function(p, model){
  q <- demand_51tax(p, model)
  cs <- 0.5*(model$coefficients[[1]] - p)*q
  return(cs)
}

CSl_51 <- CS_51tax(peq, model_demand_l)/100 - env_cost51/100
```

**The welfare of low income consumer is $7604.12**

(e) Power suppliers (i.e., electricity producers)

```{r}
supply <- function(q){
  p <- MPC_slope * q
  return(p)
}

PS <- function(q){
  0.5 * supply(q) * q
}

PS_new <- PS(qeq)/100
```

**The welfare of power suppliers is $23309.80**

(f) Total environmental damage

```{r}
env_cost51 <- (1.97 * qeq)/100
```

**The total environmental damage is $9854.26**

(g) Total tax revenue generated

```{r}
tax_rev51 <- (1.97 * qeq)/100
```

**Total tax revenue generated is $9854.26**

5. Now, assume that all revenue from the electricity tax will be redistributed to the consumers in proportion to their pre-tax consumption. For example, if 80% of the electricity was consumed by “high” income consumers, then they get 80% of the tax revenue. Additionally, consider the fact that current scientific evidence suggests the true SCC may be much higher than \$51. For a range of SCC values (\$51, \$75, \$100, \$125, and $150 per metric ton of CO2), calculate the effects of an SCC-based electricity tax on:

we need:
total q (qeq)
qhigh
qlow
- get percentage
- multiply that by (t* * qeq)
- add to CS

(a) Overall welfare of “high” income consumers
(b) Overall welfare of “low” income consumers
(c) Electricity producers
```{r}
MEC_51 <- (0.85/2204.62)*51 #1.97 cents
MEC_75 <- (0.85/2204.62)*75 #2.89 cents
MEC_100 <- (0.85/2204.62)*100 #3.86 cents
MEC_125 <- (0.85/2204.62)*125 #4.82 cents
MEC_150 <- (0.85/2204.62)*150 #5.78 cents

SCC_vec <- c(51, 75, 100, 125, 150)
MEC_vec <- c(MEC_51, MEC_75, MEC_100, MEC_125, MEC_150)

equilibrium <- function(p){
  (demand(p + i , model_demand_l) + demand(p + i,model_demand_h))-(p/MPC_slope)
  
  peq_new <- uniroot(fm_eq, c(0,750000))$root
  
  qeq_new <- demand_agg_new(peq_new)
  
  }

values <- data.frame()

for(i in MEC_vec){
  
  fm_eq <- function(p){
    (demand(p + i , model_demand_l) + demand(p + i,model_demand_h))-(p/MPC_slope)
  }
  
  demand_new <- function(p, model){
  q <- ((p + i) - model$coefficients[[1]])/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
  }
  
  CS_new <- function(p, model){
  q <- demand_new(p, model)
  cs <- 0.5*(model$coefficients[[1]] - p)*q
  return(cs)
  }
  
  demand_agg_fl <- function(p){
  q <- demand(p + i , model_demand_l) + demand(p + i, model_demand_h)
  return(q)}

  peq_new <- uniroot(fm_eq, c(0,750000))$root
  
  qeq_new <- demand_agg_fl(peq_new) + i
  
  qhigh <- demand(10, model_demand_h)
  
  qlow <- demand(10, model_demand_l)
  
  qhperc <- qhigh/(qhigh + qlow)
  
  qlperc <- qlow/(qhigh + qlow)
  
  encost_new <- i * qeq_new
  
  taxrevhigh <- encost_new * qhperc
  
  taxrevlow <- encost_new * qlperc
  
  CSlow_new <- (CS_new(peq_new, model_demand_l)/100) - (encost_new/100) + (taxrevlow/100)
  
  CShigh_new <- (CS_new(peq_new, model_demand_h)/100) + (taxrevhigh/100)
  
  PSnew <- PS(qeq_new)/100
  
  results <- cbind((i * 100), peq_new, qeq_new, encost_new, CSlow_new, CShigh_new, PSnew) %>% unlist(recursive = FALSE)
  
  values <- rbind(values, results)
}

values["SCC"] <- SCC_vec
values <- values[, c(8,1,2,3,4,5,6,7)]

values %>% 
  kable(digits = c(2, 2, 2, 2, 2, 2, 2),
        col.names = c("SCC (\\$)", "Optimal tax (cents)", "Equilibrium price (cents)", "Equilibrium quantity (kWh)", "Environmental cost ($)", "Consumer surplus (low-income)", "Consumer surplus (high-income)", "Producer surplus")) %>% 
        kable_styling(full_width = FALSE, position = "center") %>% 
  kable_classic()
```

6. Suppose the “high” income group has access to expensive home solar generation. This lowers the electricity demand curve for the “high” income group by half (vertically). Under this new demand:

(a) What is total electricity consumption?
```{r}
#demand function
demand_halved <- function(p, model){
  q <- ((2*p) - model$coefficients[[1]])/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
}

price = seq(0, 30, length.out = 100)

hpred_halved <- tibble(price = price,
                 demand = demand_halved(price, model_demand_h))

hpred_halved[hpred_halved == 0] <- NA

#aggregate demand
demand_agg_6 <- function(p){
  q <- demand_halved(p, model_demand_l) + demand_halved(p, model_demand_h)
  return(q)
}

#price vector
price = seq(0, 30, length.out = 100)

#applying the aggregate demand function to the price vector
Qagg_6 <- map(price, demand_agg_6) %>% unlist()

#making the new dataframe with Qagg and price
agg_data_6<- tibble(Qagg = Qagg_6, price = price)

agg_data_6[agg_data_6 == 0] <- NA


#plot
ggplot() +
  geom_line(agg_data, mapping = aes(x = Qagg, y = price, color = "Aggregate", lty = "Aggregate")) +
  geom_line(l_pred, mapping = aes(x = demand, y = price, color = "Low income", lty = "Low income")) +
  geom_line(h_pred, mapping = aes(x = demand, y = price, color = "High income", lty = "High income")) +
  geom_line(MPC, mapping = aes(x = demand, y = price, color = "MPC", lty = "MPC")) +
  geom_segment(mapping = aes(x = 0, y = 1.97, xend = 819713.75, yend = 1.97, color = "MEC", lty = "MEC")) +
  geom_line(agg_data_6, mapping = aes(x = Qagg, y = price, color = "Aggregate (new)", lty = "Aggregate (new)")) +
  geom_line(hpred_halved, mapping = aes(x = demand, y = price, color = "High income (new)", lty = "High income (new)")) +
  scale_color_manual(name = "Legend", values = cal_palette(name = "kelp1", n = 7, type = "continuous")) +
  scale_linetype_manual(name = "Legend", values = c("dotdash", "solid", "dashed", "solid", "dashed", "solid", "solid")) +
  labs(x = "Quantity (kWh)",
       y = "Price (cents)",
       color = "Demand Curve") + 
  theme_minimal()
```

```{r}
#finding the intersect
freemarket_6 <- function(p){
  (demand_halved(p, model_demand_l) + demand_halved(p, model_demand_h))-(p/MPC_slope)
  }
peq_6 <- uniroot(freemarket_6, c(0,750000))$root
qeq_6 <- demand_agg_6(peq_6)
```

**Total electricity consumption is 398978.07 kWh**

(b) What is the total environmental externality?
```{r}

env_dmg_6 <- (1.97 * qeq_6)/100

```

**Total environmental externality is $7,859.87**

(c) What value of the electricity tax makes the total environmental damage the same as the damage when solar panels are available to the high income group?
```{r}
#demand function
demand_p <- function(q, model){
  p <- q * model$coefficients[[2]] + model$coefficients[[1]]
  p <- ifelse(p<0,0,p)
  return(p)
}

#aggregate demand original
demand_agg_p <- function(q){
  p <- demand_p(q, model_demand_l) + demand_p(q, model_demand_h)
  return(p)
}

#demand halved function
demand_halved_p6 <- function(q, model){
  p <- (q*model$coefficients[[2]] + model$coefficients[[1]]) * 0.5
  p <- ifelse(p<0,0,p)
  return(p)
}

#new aggregate demand
demand_agg_p6 <- function(q){
  p <- demand_halved_p6(q, model_demand_l) + demand_halved_p6(q, model_demand_h)
  return(p)
}

diff <- demand_agg_p(qeq_6) - demand_agg_p6(qeq_6)
```

**Value of the tax is 5.43 cents**
